---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First unzip the **"activity.zip"** file.
```{r cache=TRUE}
unzip('activity.zip')
```

Next read the **"activity.csv"** file into a data.table using the
fread call. This data.table will serve as the base for the 
remainder of the study.
```{r}
library(data.table)
# Read the data into a data.table
activity <- fread( 'activity.csv',
        header=TRUE, sep=",", na.strings = "NA",
        colClasses=c('numeric', 'Date', 'numeric')
    )
# Define the indexing key on the data.table
setkey(activity, date)
```


## What is mean total number of steps taken per day?
```{r out.width=900}
library(ggplot2)

# 1. Calculate steps taken per day.
activity[, daily_steps:=sum(steps), by=date]

# 2. Build a histogram of the steps taken for each day.
#   Use the steps as weight for this calculation.
plot <- qplot(date, data=activity, weight=steps)
#   theming
plot + theme_minimal() + theme(axis.text.x = element_text(angle=90, size = 8))

# 3. Calculate and report the mean and median of the total number of steps taken.
# Subset the unique dates from the data table, returning the daily_steps vectors.
unique_days_steps <- unique(activity)$daily_steps

# Mean value for the number of steps observed daily.
mean(unique_days_steps, na.rm=TRUE)
# Median value for the number of steps observed daily.
median(unique_days_steps, na.rm=TRUE)
```

## What is the average daily activity pattern?
```{r out.width=900}
# Find the average by interval over all of the observed days
activity[,interval_average:=mean(steps, na.rm=TRUE),by=interval]

# Build the time series
plot <- qplot(
    interval, interval_average, data=activity,
    geom="line", xlab="5-minute interval", ylab="steps taken",
    main="Time series report on average activity over two months."
)
plot + theme_minimal()

# Obtain the most active interval period
max_row <- activity$interval_average == max(activity$interval_average)
most_active_interval <- unique(activity[max_row,]$interval)
```

The most active interval is **`r most_active_interval`**

## Imputing missing values
```{r out.width=900}
na_activity_filled <- copy(activity)
# 1. Get the number of rows with NAs
nrow(na_activity_filled[is.na(na_activity_filled$steps),])

# 2. Fill in all NA values with the 5 minute interval mean calculated above.
na_activity_filled[is.na(na_activity_filled$steps), steps:=interval_average]

# 1. Calculate steps taken per day.
na_activity_filled[, daily_steps:=sum(steps), by=date]

# 2. Build a histogram of the steps taken for each day.
#   Use the steps as weight for this calculation.
plot <- qplot(date, data=na_activity_filled, weight=steps)
#   theming
plot + theme_minimal() + theme(axis.text.x = element_text(angle=90, size = 8))

# 3. Calculate and report the mean and median of the total number of steps taken.
# Subset the unique dates from the data table, returning the daily_steps vectors.
unique_days_steps <- unique(na_activity_filled)$daily_steps

# Mean value for the number of steps observed daily.
mean(unique_days_steps, na.rm=TRUE)
# Median value for the number of steps observed daily.
median(unique_days_steps, na.rm=TRUE)
```

* **Do the calculated values change from the first part of this assignment?**
    Yes.
    
* **What is the impact of inputting missing data on the estimates of the total daily steps?**
    First and most obviously any days with **NA** present in their record has an automatic increase to their values.
    This is most visual in the first few days where they're largely **NA** data.
    
    The method used to replace the NA values also brings the median and mean in-line with one another because the
    calculation being used was the mean for that interval period.


## Are there differences in activity patterns between weekdays and weekends?
```{r out.width=900}
# Ensure all of the dates are date objects
activity$date <- as.Date(activity$date)
# Assign the week days
activity[,weekday := weekdays(date)]
# Assign weekend vs weekday
weekend_days = c('Saturday', 'Sunday')
activity[weekday == weekend_days, day_type:='weekend']
activity[weekday != weekend_days, day_type:='weekday']

# Add another column that tracks the average steps during a time interval plus day type 
#   combination.
activity[,day_type_interval_average:=mean(steps, na.rm=TRUE),by=list(interval, day_type)]

# Plot the two types (weekdays, weekends) and their respective average activites.
plot <- qplot(interval, day_type_interval_average, data=activity, facets=day_type ~ ., geom="line")
plot + theme_minimal()
```